Performance-portable Parallel Programming in Computer Science Education with Kokkos

### 2020-01-23 src/2.GenericProgramming.tex
Finished out the "triad" text

The parallel programming paradigm is an abstract ~\emph{representation} with the purpose of facilitating the programmer's understanding of programming rules and program behavior. Which programming paradigm to chose depends on several considerations. Parallel-patterns allow to express concurrency for commonly occurring programming patterns such as loop constructs. Tasking is a paradigm that support the expression of concurrent loops as well as irregular algorithms. Distributed and correctness-oriented programming models may implement actor-based programming. In this programming model, each unit of execution represent an actor who communicates over predefined communication interfaces. This paradigm eliminates accesses to shared state and align well with message passing programming (MPI). The execution model and memory model of a programming model defines behavior. That is, it defines the relationship between abstract concepts and program behavior on the given architecture.

Lastly, a model must decide between declarative and imperative semantics. Declarative semantics require a developer to express intent, implementation is the responsibility of the programming model and underlying toolchain. Imperative semantics require the developer to specify the mechanisms by which their intent will be implemented. In an imperative model, developer success is tied to their knowledge of implementation mechanisms. In a declarative model that success is tied to the developer's ability to accurately declare their requirements, and the toolchain's ability to turn that into an implementation. Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned types of semantics. While both programming models implement the same programming paradigm (parallel patterns), Figure~\ref{figOMPLike} required the developer to express how to parallelize the loop, while in~\ref{figKokkosLike} the developer just declared their parallel semantics, and relied on the programming model to pick an implementation.

>>>>

%The parallel programming paradigm is an abstract ~\emph{representation} with the purpose of facilitating the programmer's understanding of programming rules and program behavior. Which programming paradigm to chose depends on several considerations. Parallel-patterns allow to express concurrency for commonly occurring programming patterns such as loop constructs. Tasking is a paradigm that support the expression of concurrent loops as well as irregular algorithms. Distributed and correctness-oriented programming models may implement actor-based programming. In this programming model, each unit of execution represent an actor who communicates over predefined communication interfaces. This paradigm eliminates accesses to shared state and align well with message passing programming (MPI). The execution model and memory model of a programming model defines behavior. That is, it defines the relationship between abstract concepts and program behavior on the given architecture.

The paradigm most central to Kokkos are standard parallel patterns. The three main primitives provided to developers are \emph{map}, \emph{reduce}, and \emph{scan}, with some ability for composition. We found that these could pretty easily be mapped to a wide variety of architectures, and that many algorithms can be expressed in terms of these building blocks. Many software engineering curricula already cover patterns, for many students this will resonate immediately. For those that don't, this provides an excuse to discuss another important topic in the field.

Lastly, a model must divide responsibility between. In practice, this is largely a choicebetween declarative and imperative semantics. Declarative semantics require a developer to express intent, implementation is the responsibility of the programming model and underlying toolchain. Imperative semantics require the developer to specify the mechanisms by which their intent will be implemented. In an imperative model, developer success is tied to their knowledge of implementation mechanisms. In a declarative model that success is tied to the developer's ability to accurately declare their requirements, and the toolchain's ability to turn that into an implementation. Figures~\ref{figOMPLike} and~\ref{figKokkosLike} show examples of two sample applications using the aforementioned types of semantics. While both programming models implement the same programming paradigm (parallel patterns), Figure~\ref{figOMPLike} required the developer to express how to parallelize the loop, while in~\ref{figKokkosLike} the developer just declared their parallel semantics, and relied on the programming model to pick an implementation.
